Local Spectral Clustering of Density Upper Level Sets

### 2019-01-23 icml/icml_local_density_clustering.tex
Edits through Sect 1

\theoremstyle{aldenrmrk}
\newtheorem{remark}{Remark}

\newcommand{\theHalgorithm}{\arabic{algorithm}}

\icmltitlerunning{Local clustering of density upper level sets}

@@

\twocolumn[
\icmltitle{Local Spectral Clustering of Density Upper Level Sets}

\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Alden Green}{cmu}
\icmlauthor{Sivaraman Balakrishnan}{cmu}
\icmlauthor{Ryan Tibshirani}{cmu}
\end{icmlauthorlist}

\icmlaffiliation{cmu}{Department of Statistics and Data Science, Carnegie Mellon University, Pittsburgh PA, USA}

\icmlcorrespondingauthor{Alden Green}{ajgreen@andrew.cmu.edu}

\icmlkeywords{local clustering}

@@

\printAffiliationsAndNotice{} % otherwise use the standard text.

\begin{abstract}
	Spectral clustering methods are 
	a family of popular non-parametric clustering tools.
	Recent works have proposed and analyzed 
	\emph{local} spectral methods, which 
	extract clusters using
	locally-biased random walks around a user-specified seed node, and are known to have worst-case guarantees for certain graph-based measures of cluster quality.
	In contrast to existing works, 
	we analyze the personalized PageRank (\ppr) algorithm in the classical statistical learning setup,
	where we obtain samples from an unknown distribution, and aim to identify connected
	regions of high-density (density clusters).
	We introduce a bicriteria for evaluating cluster quality and provide guarantees for these criteria evaluated on empirical analogues to these density clusters. As a result, the \pprspace algorithm run over a neighborhood graph is shown to 
	extract sufficiently salient density clusters.
\end{abstract}

\section{Introduction}
\label{introduction}

Let $\mathbf{X} := (x_1, \ldots, x_n)$ be a sample drawn i.i.d from a distribution $\Pbb$ with density $f$ supported on $\Rd$. Our statistical learning task is clustering: splitting data into groups which satisfy some notion of within-group similarity and between-group difference.

Spectral clustering methods are a family of powerful non-parametric clustering algorithms. Let $G = (V,E)$ be an unweighted and undirected graph, with $\Abf \in \Reals^{V \times V}$ the symmetric adjacency where $A_{uv} = 1$ if $(u,v) \in G$. We form the random walk transition probability matrix $\Wbf$, and corresponding graph Laplacian matrix $\Lbf$\footnote{Often, either of the Laplacian matrices  $\Lbf_{sym} := \Dbf^{\frac{1}{2}}\Wbf \Dbf^{-\frac{1}{2}}$ or $\Lbf_{unn} := \Dbf - \Abf$ are used instead.}:
\begin{equation}
\label{eqn: random_walk_laplacian}
\Wbf := \Dbf^{-1}\Abf; ~~~ \Lbf = \Ibf{n} - \Wbf
\end{equation}
where the degree matrix $\Dbf$ is a diagonal matrix with $\Dbf_{uu} := \sum_{v \in V} \Abf_{uv}$, and $\Ibf{n}$ is the $n \times n$ identity matrix.

Roughly speaking, spectral clustering techniques first learn an embedding the data $\Xbf$ using the spectrum of the graph Laplacian, and subsequently 
apply a simple clustering technique (such as k-means) to this \emph{spectral embedding}.
When applied to large graphs (or large point clouds) classical global spectral 
methods can be computationally cumbersome and 
insensitive to the local geometry of the distribution of the samples
\citep{mahoney2012,leskovec2010}.
This in turn has led to the investigation of local spectral algorithms \citep{spielman2013,anderson2006,leskovec2010}
which leverage locally-biased spectra computed using random walks around 
a user-specified seed node.

We are interested in understanding what the output of a clustering algorithm on $\Xbf$ reveals about the unknown density $f$. For $\lambda > 0$ and the upper level set $\set{x: f(x) \geq \lambda}$, it is intuitive \cite{hartigan1981,chaudhuri2010} to define clusters as the connected components $\mathbb{C}_f(\lambda)$ of the upper level set; we call these connected regions of high density \textit{density clusters}, and study the ability of spectral methods to identify such clusters.

\paragraph{Graph connectivity criteria.}

A somewhat more standard mode of understanding spectral clustering methods is to view them as approximating some graph connectivity criteria. There are many graph-based measures which assess the cluster quality of a subset $S \subseteq V$ (or more generally the quality of a partition $S_1 \cup \ldots \cup S_m = V$, for $m \geq 2$. See \cite{yang2015,fortunato2010} for an overview.)

Arguably a natural way to assess cluster quality is via a pair of criteria capturing the \emph{external} and \emph{internal connectivity} of $S$, respectively.  We define the external connectivity of a subset $S \subseteq V$ to be its normalized cut. For $S' \subset V$, the cut between $S$ and $S'$ is
\begin{equation*}
\cut(S, S'; G) := \sum_{u \in S} \sum_{v \in S'} \1\left((u,v) \in E \right)
\end{equation*}
and the volume of $S$ is
\begin{equation*}
\vol(S; G) := \sum_{u \in S} \sum_{v \in V} \1\left((u,v) \in E \right).
\end{equation*}
Then, the \emph{normalized cut} of $S$ is
\begin{equation}
\label{eqn: norm_cut}
\Phi(S; G) := \frac{\cut(S; G)}{ \min\left\{\vol(S; G), \vol(S^c; G) \right\} }
\end{equation} 
(where we abbreviate $\cut(S; G) = \cut(S; S^c; G)$).

Given $S \subseteq V$, the subgraph induced by $S$ is given by $G[S] = (S, E_S)$, where $(u,v) \in E_S$ if both $u$ and $v$ are in $S$ and $(u,v) \in E_S$.
Our internal connectivity parameter $\Psi(S)$ will capture the time it takes for the random walk over $G[S]$ to mix (that is, approach a stationary distribution) uniformly over $S$. Letting $\abs{S} = m$, $\Abf_S$ denotes the $m \times m$ adjacency matrix representation of $G[S]$; similarly, $\Dbf_S$ is the diagonal degree matrix with entries $(\Dbf_{S})_{ii} = \sum_{j: v_j \in S} \Abf_{ij}$, and $\Wbf = \Dbf_S^{-1} \Abf_S$ is the corresponding random walk matrix (again, defined only over $G[S]$.)

For $v \in V$ we write
\begin{equation}
\label{eqn: random_walk}
\pibf = (\pi_{u})_{u \in S}, ~~ q_{vu}^{(m)} = e_v\Wbf_S^m e_u
\end{equation}
for the stationary distribution and the $m$-step transition probability of the random walk over $G[S]$ originating at $v$.
\footnote{Given a starting node $v$ and and a random walk defined by transition probability matrix $\mathbf{P}$, the rotation $e_v \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$ steps. }
(Of course, given the definition of $\Wbf_S$ it is of course well known that the stationary distribution $\widetilde{\pibf}$ is given by $\widetilde{\pi}_{u} = (\Dbf_{S})_{uu} / \vol(S; G[S])$.)

The \emph{relative pointwise mixing time} is defined as 
\begin{equation*}
\tau_{\infty}(\qbf; G[S]) := \min\set{m: \forall u,v \in V, \frac{\abs{q_{vu}^{(m)} - \pi_u}}{\pi_u} \leq 1/4} 
\end{equation*}
where $\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}, \qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$.

The internal connectivity parameter $\Psi(S; G)$ is simply one over the mixing time:
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(\qbf; G[S])}
\end{equation}

The graph clustering task is thus find a subset $S$ (or, for global algorithms, a partition $S_1 \cup \ldots \cup S_m = V$), which has both small external and large internal connectivity. If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no less than $\Psi$, we will refer to it as a $(\Phi,\Psi)$-cluster. Both local \cite{zhu2013} and global \cite{kannan04} spectral algorithms have been shown to output clusters (or partitions) which provably satisfy approximations to the optimal $(\Phi, \Psi)$-cluster (or partition) for a given graph $G$. \footnote{In the case of \cite{kannan04}, the internal connectivity parameter $\phi$ is actually the conductance, i.e. the minimum normalized cut within the subgraph $G[S]$. See Theorem 3.1 for details; however, note that $\phi^2 / \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a lower bound on $\Psi$.}

\paragraph{Personalized PageRank.}
As mentioned previously, global algorithms which find spectral cuts may be computationally infeasible for large graphs; in this setting, local algorithms may be preferred or even required. We will restrict our attention in particular to one such popular algorithm: \emph{personalized PageRank} (\ppr). The personalized PageRank algorithm was first introduced by \cite{haveliwala2003} and variants of this algorithm have been studied further in several recent works \citep{spielman2011,spielman2014,zhu2013,anderson2006,mahoney2012}.

The random walk matrix $\Wbf$ over the graph $G = (V,E)$ with associated adjacency matrix $\Abf$ is given by (\ref{eqn: random_walk_laplacian}). \pprspace is then defined with respect to the following inputs: a user-specified seed node $v \in V$ , and  $\alpha \in [0,1]$ a teleportation parameter. Letting $e_{v}$ be the indicator vector for $v$ (meaning $e_v$ has a $1$ in the $v$th location and $0$ everywhere else), the \emph{PPR vector} is the solution to the following linear equation:
\begin{equation}
\label{eqn: ppr_vector}
\pbf(v,\alpha;G) := \alpha \ebf{v} + (1 - \alpha) \pbf(v,\alpha;G) \Wbf
\end{equation}
We note in passing that, for $\alpha > 0$, the vector $\pbf(v,\alpha;G)$ can be well-approximated by a simple local computation (of a random walk with restarts at the node $v$.) We also point out that, from a density clustering standpoint, since density clusters are inherently local, using the PPR algorithm eases the analysis, and as we will observe in the sequel our analysis requires fewer global regularity conditions relative to more classical global spectral algorithms.

To compute a cluster $\widehat{C} \subset V$ using the \pprspace vector, we will take sweep cuts of $\pbf(v, \alpha; G)$. To do so we require another parameter $\pi_0$. For a number $\beta \geq 0$, the sweep cut $S_\beta$ is then
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \set{u \in V: p_u > \beta \pi_{0}}.
\end{equation}
We will choose one such sweep cut to be our cluster estimate $\widehat{C}$.

\paragraph{Neighborhood graph.}

To formally introduce the local clustering algorithm we consider, we must first give a method for forming a graph over the data $\Xbf$.

Let $(r_n)$ be a sequence of positive numbers. Given a sequence of kernel functions $k_n: \Rd \times \Rd \to [0,\infty)$ of the form $k_n(x,x') = k(\norm{x - x'}/ r_n)$ for $k$ a non-increasing function, and data $\Xbf = \set{x_1, \ldots, x_n}$ sampled from $\Pbb$ as before, form the \emph{neighborhood graph} $G_n = (\Xbf, E_n)$ with $E_n = \set{k(x_i,x_j): 1 \leq i < j \leq n}$. (Here, $\norm{\cdot}$ is used to denote Euclidean norm).

Algorithm \ref{alg: ppr} will be the simple variant of PPR we analyze. It will take as input the data $\Xbf$ along with user-specified parameters $r, \alpha$, $\pi_0$, and $v \in \Xbf$. 
\begin{algorithm}
	\caption{PPR on a neighborhood graph}
	\label{alg: ppr}	
	{\bfseries Input:} data $\Xbf$, radius $r$, teleportation parameter $\alpha \in [0,1]$, seed node $v \in \Xbf$, target $\pi_0$. \\
	{\bfseries Output:} $\widehat{C} \subset V$.
	\begin{algorithmic}[1]
		\STATE Form the neighborhood graph $G_{n,r}$ as given in (\ref{eqn: neighborhood_graph})
		\STATE Compute PPR vector $\pbf(v, \alpha; G_{n,r})$  as defined by (\ref{eqn: ppr_vector}).
		\STATE For $\beta \in (\frac{3}{10}, \frac{1}{2})$ compute sweep cuts $S_{\beta}$ as defined by (\ref{eqn: sweep_cuts}).
		\STATE Return
		\begin{equation*}
		\label{eqn: sweep_cuts_min_conductance}
		\widehat{C} = \argmin_{\beta \in (\frac{3}{10}, \frac{1}{2})} \Phi(S_{\beta}; G_{n,r})
		\end{equation*}
	\end{algorithmic}
\end{algorithm}

We note that the choice of a specific interval such as $(\frac{3}{10}, \frac{1}{2})$ for the range of $\beta$ is standard in the analysis of \pprspace algorithms.

It is worth calling attention to some other work on computing the normalized cut over neighborhood graphs. In this context, continuous analogues to (for instance) normalized cut have been defined, over the data-manifold rather than the graph, and convergence finite sample graph-theoretic functionals to their continuous counterparts has been shown
\cite{garciatrillos16, arias-castro12, maier11}.
However, in addition to the graph-minimization problem being computationally infeasible, these continuous analogues are not always easily interpretable -- and their corresponding minimizers not always easily identifiable -- for the particular density function under consideration. Of course, relating these partitions to the arguably more simply defined high density clusters can be also challenging in general. Intuitively, however, under the right conditions such high-density clusters should have more edges within themselves than to the remainder of the graph. We formalize this intuition next.

%It is worth pointing out that in this context, some theory has been developed regarding how graph theoretic quantities such as the normalized cut $\Phi$ (and others) relate to properties of the underlying distribution $f$ as well as the kernel function $k$. Such analyses typically proceed by defining a continuous analogue to the measure of cluster quality under consideration. Then, under appropriate specification of $k$ and a proper schedule of $(r_n)$, convergence of clusters output by spectral (and other) algorithms to the corresponding minima of these continuous analogues has been shown \cite{vonluxburg2008, garciatrillos18}.

\subsection{Summary of results}

Hereafter, we consider the uniform kernel function for a fixed $r > 0$,
\begin{equation}
k(x,x') = \1(\norm{x - x'} \leq r)
\end{equation}
and the associated neighborhood graph
\begin{equation}
\label{eqn: neighborhood_graph}
G_{n,r} = (\Xbf, E_{n,r}), \text{  $(x_i,x_j) \in E_{n,r}$ if $k(x_i,x_j) = 1$}
\end{equation}
For a given high density cluster $\Cset \subseteq \Cbb_f(\lambda)$, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. We now introduce a notion of consistency for the task of density cluster estimation:

\begin{definition}[Consistent density cluster estimation]
	\label{def: consistent_density_cluster_estimation}
	For an estimator $\widehat{\Cset}_n \subset \Xbf$, and any $\Cset, \Cset' \in \Cbb_f(\lambda),$ we say $\widehat{\Cset}_n$ is a consistent estimator of $\Cset$ if the following statement holds: as the sample size $n \to \infty$, each of the following
	\begin{equation}
	\label{eqn: consistent_density_cluster_recovery}
	\Cset[\Xbf] \subseteq \widehat{\Cset}_n, ~\mathrm{ and }~ \widehat{\Cset}_n \cap \Cset'[\Xbf] = \emptyset
	\end{equation}
	occur with probability tending to $1$.

\end{definition}

Our results can now be summarized by the following points:

\begin{enumerate}
	\item 
	Under a natural set of geometric conditions
	\footnote{We formally introduce the geometric conditions in Section \ref{sec: background}. They preclude clusters which are too thin and long, or those for which the gap in density between the high density area and the outside is not sufficiently large}, 
	the normalized cut of an empirical density cluster $C[\Xbf]$ can be bounded. Theorem \ref{thm: conductance_upper_bound} provides an upper bound.

	\item Under largely the same set of geometric conditions to Theorem \ref{thm: conductance_upper_bound}, Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex} provides a lower bound on the inverse mixing time of a random walk over $C[\Xbf]$. Theorem \ref{thm: inverse_mixing_time_lower_bound} gives a tighter lower bound, but requires the additional (restrictive) assumption of convexity.

	\item 
	In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we show these bounds on the graph connectivity criteria have algorithmic consequences for personalized PageRank. we show that a careful analysis of the form typical to local clustering algorithms yields Theorem \ref{thm: consistent_recovery_of_density_clusters}, which states that Algorithm \ref{alg: ppr}, properly initialized, performs consistent density cluster estimation in the sense of (\ref{eqn: consistent_density_cluster_recovery}).

	\item Corollary \ref{cor: ppr_cluster} follows as an immediate consequence of Theorems \ref{thm: conductance_upper_bound} and \ref{thm: inverse_mixing_time_lower_bound}, along with the previous work of \cite{zhu2013}. It gives an upper bound on the normalized cut of the set $\widehat{C}$ output by Algorithm \ref{alg: ppr}, as well as upper bounding the symmetric set difference between $\widehat{C}$ and $\Cset[\Xbf]$.
\end{enumerate}

\paragraph{Organization.}
In Section \ref{section: examples}, we provide some example density functions, to clarify the relevance of our results, and empirically demonstrate that violations of the geometric conditions we set out in Section \ref{sec: background} manifestly impact density cluster recovery (i.e. the conditions are not superfluous), before concluding in \ref{sec: discussion}. First, however, we summarize some related work.

\subsection{Related Work}
In addition to the background given above, a few related lines of work are worth highlighting.

>>>>

\theoremstyle{aldenrmrk}
\newtheorem{remark}{Remark}
\newcommand{\theHalgorithm}{\arabic{algorithm}}

%%%%%
\icmltitlerunning{Local Spectral Clustering of Density Upper Level Sets}

@@

\twocolumn[
\icmltitle{Local Spectral Clustering of Density Upper Level Sets}
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Alden Green}{cmu}
\icmlauthor{Sivaraman Balakrishnan}{cmu}
\icmlauthor{Ryan J. Tibshirani}{cmu}
\end{icmlauthorlist}

\icmlaffiliation{cmu}{Department of Statistics and Data Science, Carnegie Mellon
  University, Pittsburgh PA, USA} 
\icmlcorrespondingauthor{Alden Green}{ajgreen@andrew.cmu.edu}
\icmlkeywords{local clustering}

@@

\printAffiliationsAndNotice{}

\begin{abstract}
\vskip 0.1 in % RJT: there was no space here and it looked very squished 
Spectral clustering methods are a family of popular nonparametric clustering
tools.  Recent works have proposed and analyzed \emph{local} spectral methods,
which extract clusters using locally-biased random walks around a user-specified
seed node.  Several local methods, such as the personalized PageRank (PPR)
algorithm, have been shown to have worst-case guarantees for certain graph-based
measures of cluster quality.  In contrast to existing works, we analyze PPR in a
traditional statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We introduce two natural criteria for cluster quality, and derive
bounds for these criteria when evaluated on empirical analogues of density
clusters. Moreover, we prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters.   
\end{abstract}

\section{Introduction}
\label{sec: introduction}

Let $\mathbf{X} = \{x_1, \ldots, x_n\}$ be a sample drawn i.i.d.\ from a
distribution $\Pbb$ on $\Rd$, with density $f$, and consider the problem of 
clustering: splitting the data into groups which satisfy some notion of
within-group similarity and between-group difference.  We focus on spectral
clustering methods, a family of powerful nonparametric clustering algorithms.
Roughly speaking, a spectral technique first constructs a geometric graph $G$,
where vertices are associated with samples, and edges correspond to proximities
between samples. It then learns a feature embedding based on the Laplacian of
$G$, and applies a simple clustering technique (such as k-means clustering) in
the embedded feature space.

To be more precise, let $G=(V,E,w)$ denote a weighted, undirected graph  
constructed from the samples $\mathbf{X}$, where $V=\{1,\ldots,n\}$, and $w_{uv}
= K(x_u,x_v) \geq 0$ for $u,v \in V$, and a particular kernel function $K$.
Here $(u,v) \in E$ if and only if $w_{uv} > 0$.  We denote by $\Abf \in
\Reals^{n \times n}$ the weighted adjacency matrix, which has entries
$A_{uv}=w_{uv}$, and by $\Dbf$ the degree matrix, with 
\smash{$\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$}.  We also denote by $\Wbf,\Lbf$
the random walk transition probability matrix and normalized\footnote{Other
  popular choices here include the unnormalized Laplacian, and symmetric
  normalized Laplacian.} 
Laplacian matrix, respectively, which are defined as
$$
\Wbf = \Dbf^{-1}\Abf, \quad \Lbf = \Ibf{} - \Wbf,
$$
where $\Ibf{} \in \Reals^{n\times n}$ is the identity matrix.  Classical global
spectral methods take a eigendecomposition $L=U \Sigma U^T$, use some 
number of eigenvectors (columns in $U$) as a feature representation for the
samples, and then run (say) k-means in this new feature space.

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be been computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citet{haveliwala2003}, and further developed by
\citet{spielman2011,spielman2014,anderson2006,mahoney2012,zhu2013}, among 
others.

Local spectral clustering techinques have been practically very successful
\citep{leskovec2010,mahoney2012,REFS}, which
has led many authors to develop theory that explains their success 
\citep{zhu2013,REFS},
from the perspective of worst-case guarantees on traditional graph-theoretic
notions of cluster quality (like conductance).  In this paper, we adopt a more
traditional statistical viewpoint, and examine what the output of a local
clustering algorithm on $\Xbf$ reveals about the unknown density $f$.  
In particular, we examine the ability of the PPR algorithm to recover
\emph{density clusters} of $f$, which are defined as the connected components of
the upper level set $\{x : f(x) \geq \lambda\}$ for some threshold $\lambda >
0$ (a central object of central interest in the classical statistical literature
on clustering, dating back to \citealt{hartigan1981}).

\subsection{Graph Connectivity Criteria}

Here we define a pair of criteria that reflect the quality of a cluster with
respect to $G=(V,E,w)$.  There are many graph-based measures of cluster quality  
that one could consider; see, e.g., \citet{yang2015,fortunato2010} for 
an overview.  The pair of criteria that we focus on are (arguably) quite
natural, and moreover, they play a fundamental role in our analysis 
of the PPR algorithm.  Our two critera capture the \emph{external} and
\emph{internal} connectivity of a subset $S \subseteq V$, denoted $\Phi(S; G)$
and $\Psi(S; G)$, respectively, and defined below in turn.

\paragraph{External Connectivity: Normalized Cut.}

Define the cut between subsets $S,S' \subseteq V$ to be
$$
\cut(S, S'; G) = \sum_{u \in S} \sum_{v \in S'} w_{uv},
$$
and define
\smash{$\vol(S; G) = \cut(S, V; G) = \sum_{u \in S} \sum_{v \in V} w_{uv}$}.  
As our notion of external connectivity, we use the \emph{normalized cut}  
of $S$, defined as 
\begin{equation}
\label{eqn: norm_cut}
\Phi(S; G) = \frac{\cut(S; G)}{\min\{\vol(S; G), \vol(S^c; G)\}},
\end{equation} 
where we abbreviate $\cut(S; G) = \cut(S; S^c; G)$.

\paragraph{Internal Connectivity: Inverse Mixing Time.}

For $S \subseteq V$, denote by $G[S] = (S, E_S, w_S)$ the subgraph induced by 
$S$ (where the edges are $E_S = E \cap (S \times S)$). Let $\Abf_S,\Dbf_S$
be the adjacency matrix and degree matrix, respectively, of $G[S]$.  Define the
random walk matrix as usual, $\Wbf = \Dbf_S^{-1} \Abf_S$, and for $v \in V$,
write 
\begin{equation}
\label{eqn: random_walk}
q_{vu}^{(t)} = e_v\Wbf_S^t e_u
\end{equation}
for the $t$-step transition probability of a random walk over $G[S]$
originating at $v$.\footnote{Given a starting node $v$ and and a random walk
  defined by transition probability matrix $\mathbf{P}$, the rotation $e_v
  \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$
  steps.}   Also write \smash{$\tilde\pibf = (\tilde\pi_{u})_{u \in S}$} for the 
  stationary distribution of this random walk.  (Given the definition of
  $\Wbf_S$, it is well-known that the stationariy distribution is given by
  \smash{$\tilde\pi_{u} = (\Dbf_{S})_{uu} / \vol(S; G[S])$}.)

Our internal connectivity parameter will capture the time it takes for the
random walk over $G[S]$ to mix (approach the stationary distribution)
uniformly over $S$.  For this, we first define \emph{relative pointwise mixing
  time} of $G[S]$ as 
\begin{equation*}
\tau_{\infty}(\qbf; G[S]) = \min\set{ t: \frac{|q_{vu}^{(m)} -
      \tilde\pi_u|}{\tilde\pi_u} \leq \frac{1}{4}, \; \text{for $u,v \in V$}}, 
\end{equation*}
where \smash{$\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}$}, and
\smash{$\qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$}. Now our internal 
connectivity parameter is simply the inverse mixing time,
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(\qbf; G[S])}.
\end{equation}

If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no
less than $\Psi$, we call it as a \emph{$(\Phi,\Psi)$-cluster}. Both
local \cite{zhu2013} and global \cite{kannan04} spectral algorithms have been
shown to output clusters (or partitions) which approximate the optimal $(\Phi,
\Psi)$-cluster (or partition) for a given graph $G$.\footnote{In the case of
  \cite{kannan04}, the internal connectivity parameter $\phi$ is actually the
  conductance, i.e., the minimum normalized cut within the subgraph $G[S]$. See
  Theorem 3.1 in their paper for details; however, note that $\phi^2 /
  \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a
  lower bound on $\Psi$.}

\subsection{PPR on a Neighborhood Graph}

We now describe the clustering algorithm that will be our focus for the rest of 
the paper. We start with the geometric graph that we form based on the samples 
$\Xbf$: for a radius $r > 0$, we consider the \emph{$r$-neighborhood graph} of 
$\Xbf$, denoted $G_{n,r}=(V,E)$, an unweighted graph with vertices
$V=\{1,\ldots,n\}$, and an edge $(u,v) \in E$ if and only if $\norm{x_u - x_v}
\leq  r$, where $\norm{\cdot}$ denotes Euclidean norm.  Note that this is a
special case of the general construction introduced above, with 
$K(u,v) = 1(\norm{x_u - x_v} \leq r)$.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, with respect to  
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf$ is the random walk matrix of the underlying graph $G_{n,r}$ 
and $e_{v}$ denotes indicator vector for node $v$ (with a 1 in the $v$th
position and 0 elsewhere).  In practice, we can approximately solve the above
linear system via a simple, efficient random walk, with appropriate restarts to
$v$.

For a level $\beta > 0$ and a target volume $\pi_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: p_u > \beta \pi_{0}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{3}{10},\frac{1}{2})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{3}{10}, \frac{1}{2})$} is standard in the analysis of \pprspace
algorithms, see, e.g., \citet{zhu2013}.}
we output a cluster \smash{$\widehat{C} = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a Neighborhood Graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation 
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target volume $\pi_0 >
0$. \\   
{\bfseries Output:} cluster $\widehat{C} \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn:
    ppr_vector}. 
  \STATE For $\beta \in (\frac{3}{10}, \frac{1}{2})$ compute sweep cuts
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return $\widehat{C} = S_{\beta^*}$, where 
  $$
  \beta^* = \argmin_{\beta \in (\frac{3}{10}, \frac{1}{2})} \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

\subsection{Summary of Results}

% It is worth calling attention to some other work on computing the normalized
% cut over neighborhood graphs. In this context, continuous analogues to (for
% instance) normalized cut have been defined, over the data-manifold rather than
% the graph, and convergence finite sample graph-theoretic functionals to their
% continuous counterparts has been shown 
% \cite{garciatrillos16, arias-castro12, maier11}. 
% However, in addition to the graph-minimization problem being computationally
% infeasible, these continuous analogues are not always easily interpretable --
% and their corresponding minimizers not always easily identifiable -- for the
% particular density function under consideration. Of course, relating these
% partitions to the arguably more simply defined high density clusters can be
% also challenging in general. Intuitively, however, under the right conditions
% such high-density clusters should have more edges within themselves than to
% the remainder of the graph. We formalize this intuition next.

%% RJT: I didn't know where to put this.  It was out of place, and now I'm not
%% sure where it goes ... should it go in related work?

%It is worth pointing out that in this context, some theory has been developed
%regarding how graph theoretic quantities such as the normalized cut $\Phi$ (and
%others) relate to properties of the underlying distribution $f$ as well as the
%kernel function $k$. Such analyses typically proceed by defining a continuous
%analogue to the measure of cluster quality under consideration. Then, under
%appropriate specification of $k$ and a proper schedule of $(r_n)$, convergence
%of clusters output by spectral (and other) algorithms to the corresponding
%minima of these continuous analogues has been shown \cite{vonluxburg2008,
%garciatrillos18}.

%%% RJT: This was already commented out before

Let $\Cbb_f(\lambda)$ denote the connected components of the density upper level
set $\{x : f(x) > \lambda\}$.  For a given density cluster $\Cset \in
\Cbb_f(\lambda)$, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical
  density cluster}. Below we define a notion of consistency in density cluster
estimation.

\begin{definition}[Consistent density cluster estimation]
\label{def: consistent_density_cluster_estimation}
For an estimator \smash{$\widehat{\Cset} \subseteq \Xbf$} and cluster 
$\Cset \in \Cbb_f(\lambda)$, we say \smash{$\widehat{\Cset}$} is a consistent
estimator of $\Cset$ if for all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset \not=
\Cset'$ the following holds as $n \to \infty$: 
\begin{equation}
\label{eqn: consistent_density_cluster_recovery}
\Cset[\Xbf] \subseteq \widehat{\Cset} \quad \text{and} \quad
\widehat{\Cset} \cap \Cset'[\Xbf] = \emptyset,
\end{equation}
with probability tending to 1.
\end{definition}

A summary of our main results (and outline for the rest of this paper) is as
follows.

\begin{enumerate}
\item In Section \ref{sec: PhiPsi}, we derive in Theorem \ref{thm:
    conductance_upper_bound} an upper bound on the normalized cut of an
  empirical density cluster $\Cset[\Xbf]$, under a natural set of    
  geometric conditions (precluding clusters which are too thin and long).

\item Under largely the same set of geometric conditions, we derive in Theorem
  \ref{thm: inverse_mixing_time_lower_bound_nonconvex} a lower bound on the
  inverse mixing time of a random walk over $\Cset[\Xbf]$.  We also provide in
  Theorem \ref{thm: inverse_mixing_time_lower_bound} a tighter lower bound, but
  under more restrictive assumptions (convexity of $\Cset[\Xbf]$).

\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we show in
  Theorem \ref{thm: consistent_recovery_of_density_clusters} that these bounds
  in bounds in Theorems \ref{thm: conductance_upper_bound} and \ref{thm:
    inverse_mixing_time_lower_bound_nonconvex},
  on the cluster quality criteria, have algorithmic consequences for PPR:
  properly initialized, Algorithm \ref{alg: ppr} performs consistent density
  cluster estimation in the sense of \eqref{eqn:
    consistent_density_cluster_recovery}.

\item We show in Corollary \ref{cor: ppr_cluster} that Theorems \ref{thm:
    conductance_upper_bound} and \ref{thm:
    inverse_mixing_time_lower_bound_nonconvex}, along with the results in 
  \citet{zhu2013}, lead to alternative, graph-theoretic guarantees on cluster 
  quality: an upper bound on the normalized cut of the estimated cluster 
  \smash{$\widehat{C}$}, and an upper bound on volume of the symmetric set
  difference between \smash{$\widehat{C}$} and $\Cset[\Xbf]$.

\item In Section \ref{section: examples}, we discuss the implications of our
  results for some example density functions of interest, and empirically
  demonstrate that violations of the gometric conditions we require manifestly
  impact density cluster recovery. 
\end{enumerate}

On the topic of conditions, it is worth mentioning that, as density clusters
are inherently local, focusing on the PPR algorithm actually eases our analysis
and allows us to require fewer global regularity conditions relative to those
needed for more classical global spectral algorithms.

\subsection{Related Work}

In addition to the background given above, a few related lines of work are worth
highlighting. Global spectral clustering methods were first developed in the
context of graph partitioning \citep{fiedler1973,donath1973} and their
performance is well-understood in this context (see, e.g.,
\citealt{tolliver2006,luxburg2007}).  In a similar vein, several recent works
\citep{mcsherry2001,rohe2011,kamalika2012,balakrishnan2011,lei2015,abbe2018} 
have studied the efficacy of spectral methods in successfully recovering the
community structure in the stochastic block model and variants.
